---
title: Competition Analytics Report
author: 
  - name: Dillon Labonte
    email: dillon.labonte@snhu.edu
    affiliations: 
      - name: Southern New Hampshire University
date: today
date-format: long
theme: lux
toc: true
code-fold: true
warning: false
message: false
---

```{r setup}
# including necessary libraries
library(tidyverse)
library(tidymodels)
library(patchwork)
library(kableExtra)
library(ggmosaic)

# styling
options(kable_styling_bootstrap_options = c("hover", "striped"))
theme_set(theme_bw(base_size = 14))

# read in data
data <- read_csv("data.csv")

# split data into training, test, and validation sets
set.seed(1234567890)
data_splits <- initial_split(data, prop = 0.75)

train <- training(data_splits)
test <- testing(data_splits)

```

## Statement of Purpose

This report will build and analyze classification models to predict the price range of a home based on several predictor variables. Real estate sellers may use this information to accurately price newly listed homes, which will help to limit list price errors such as being too expensive for the home characteristics, or too cheap for the home characteristics. By improving pricing strategies, this analysis helps sellers maximize their market efficiency and attract the right buyers.

## Executive Summary



## Introduction



## Exploratory Data Analysis

This exploratory data analysis aims to gives us a sense of what we are working with. It will explore properties of our response variable, `priceRange`, and some of the available predictor variables. The goal is to pick out relationships between predictor and response variables, and potentially find variables that have no impact on our response.

First we want to look at the distribution of the `priceRange` variable, which will be our response variable for our classification models.

```{r}

train |> 
  ggplot() +
  geom_bar(aes(x = priceRange), fill = "steelblue") +
  labs(title = "Distribution of Home Price Ranges",
       x = "Price Range",
       y = "Count")

```

We see that the `priceRange` response variable is relatively uniformly distributed, with a dip for the lowest price range. However, this price range is not rare which will be beneficial for our model construction, as we will be able to accurately represent all price ranges in our model. 

Let's look at some plots between our predictor variables and the `priceRange` response variable and see if we can spot any trends. First, we look at different home types and how our price range is affected.

```{r}

train |> 
  ggplot() +
  geom_bar(aes(x = homeType, fill = priceRange)) +
  labs(
    title = "Home Price Ranges due to Home Type",
    x = "Home Type",
    y = "Count",
    fill = "Price Range"
    ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

From this plot, we see that an overwhelming majority of the homes in the data set are single family homes. In this specific category, the bar plot indicates that the `priceRange` variable is relatively uniform.

Let's look at our response variable as explained by `yearBuilt`.

```{r}

train |>
  ggplot() +
  geom_boxplot(aes(y = yearBuilt, x = priceRange)) +
  labs(
    title = "Home Price Ranges due to Year Built",
    y = "Year Built",
    x = "Price Range"
  )

```

This plot indicates that the `yearBuilt` predictor variable does not have a large influence on our `priceRange` response variable, since the IQR of the year built for each price range is relatively the same.

Let's look at the lot size of each home and how it affects the price range.

```{r}

train |>
  ggplot() +
  geom_boxplot(aes(x = priceRange, y = log10(lotSizeSqFt))) +
  labs(
    title = "Distribution of Home Price Range due to Lot Size",
    x = "Price Range",
    y = "Lot Size (sq. ft.) (log scale)"
  )

```

This plot uses a log scale for the `lotSizeSqFt` since it contains some extreme outliers. In general, we see what is expected, in that the increase of lot size generally increases the price range of a home.

Let's see if schools have an impact on the price range of homes. We have a couple variables that explore this idea: `avgSchoolRating` and `MedianStudentsPerTeacher`.

```{r}
#| fig-height: 10
#| fig-width: 8

plot1 <- train |>
  ggplot() +
  geom_boxplot(aes(x = priceRange, y = avgSchoolRating)) +
  labs(
    title = "Home Price Ranges due to Average School Rating",
    x = "Price Range",
    y = "Average School Rating"
  )

plot2 <- train |>
  ggplot() +
  geom_boxplot(aes(x = priceRange, y = MedianStudentsPerTeacher)) +
  labs(
    title = "Home Price Ranges due to Median Students per Teacher",
    x = "Price Range",
    y = "Median Students per Teacher"
  )

plot1 / plot2

```

These two variables related to schools show some obvious positive trends with the `priceRange` response variable. It was expected that as `avgSchoolRating` increased, so would `priceRange`, however it is important to note that as `MedianStudentsPerTeacher` increased, we also see an increase in `priceRange`. This might battle your intuition, but the data supports this conclusion.

Let's look at some characteristics specific to our homes, the `numOfBathrooms` and `numOfBedrooms` variables.

```{r}
#| fig-height: 10
#| fig-width: 8

plot3 <- train |>
  ggplot() +
  geom_bar(aes(fill = priceRange, x = as.factor(numOfBathrooms))) +
  labs(
    title = "Home Price Ranges due to Number of Bathrooms",
    x = "Number of Bathrooms",
    y = "Count",
    fill = "Price Range"
  )

plot4 <- train |>
  ggplot() +
  geom_bar(aes(fill = priceRange, x = as.factor(numOfBedrooms))) +
  labs(
    title = "Home Price Ranges due to Number of Bedrooms",
    x = "Number of Bedrooms",
    y = "Count",
    fill = "Price Range"
  )

plot3 / plot4

```

These plots display some interesting information about our homes. In both plots, we see that as the number of bathrooms/bedrooms increase, we start to see a domination from the higher price range homes. This makes intuitive sense. On the flip side, for lower numbers of bathrooms/bedrooms we see more domination by the lower price range homes. Again, this makes intuitive sense.

This exploratory data analysis helps us to visualize our data and spot initial trends. Through this analysis, we have learned that our response variable, `priceRange`, is relatively uniformly distributed which is a desirable quality for model construction. We see that a majority of our homes in the data set are single family, and that lot size has a clear relationship with price range. We also saw that schools in the area of the homes, as well as number of bathrooms and bedrooms have clear relationships with price range. Lastly, we also saw a variable that didn't have much of an effect on our price range response, that being `yearBuilt`. This information will help us construct more effective classification models in the next section.

## Model Construction



## Model Assessment



## Conclusion



## References

